The effects of COVID-19 related social media use on well-being

### 2022-07-25 manuscript.tex
- major updates for revision - included multiple imputation - changed theory - include new data-set - extended discussion

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man,mask]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Analyzing the effects of COVID-19 related social media use on well-being},
  pdflang={en-EN},
  pdfkeywords={COVID-19, well-being, social media, news use, panel study.},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\linenumbers
\usepackage{csquotes}
\usepackage{endnotes}
\setlength{\parskip}{0em}
\raggedbottom
\note{\clearpage}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Analyzing the effects of COVID-19 related social media use on well-being}
\author{Name blinded for review\textsuperscript{1}}
\date{}

@@

\abstract{%
In times of crisis such as the Corona pandemic citizens need to stay informed about recent events, political decisions, or mandatory protection measures. To this end, many people use various types of media, and increasingly social media. However, because social media are particularly engaging, some find it hard to disconnect. In this preregistered study, I investigated whether using social media for COVID-19 related reasons affected psychological well-being. To answer this question I analyzed data from the Austrian Corona Panel Project, which consists of 3,018 participants. Well-being was measured at all 24 waves, and communication at five specific waves. I ran three random effects within between models, controlling for several stable and varying confounders. Results showed that the effects of COVID-19 related social media use on well-being were very small, arguably too small to matter. Fears that social media use during times of crisis critically impairs well-being are not supported.
}

@@

Some people actually couldn't stop using social media to learn about COVID-19 related news.
A new phenomenon emerged, termed ``doomscrolling'':
Users were glued to their screens and found it hard to pursue other relevant activities such as working, taking a break, or looking after their children (Klein, 2021).
It was asked whether using social media for COVID-19 related reasons is helpful, or whether it creates an additional burden on mental health (Sandstrom, Buchanan, Aknin, \& Lotun, 2021).
These concerns seem justified:
A study with 6,233 people from Germany conducted during the pandemic found that ``{[}f{]}requency, duration and diversity of media exposure were positively associated with more symptoms of depression'' (Bendau et al., 2021, p. 283).

As a result, with this study I want to build on this research and investigate whether or not COVID-19 related social media use affected well-being during the pandemic.
To this end, I analyzed a large-scale panel study from the Austrian Corona Panel Project (Kittel et al., 2020).
The panel consists of 24 waves and has an overall sample size of 3,018 participants.
The panel study collected a large number of psychological and demographic variables.
I explicitly aimed to investigate the causal effects of COVID-19 related social media use on well-being.

@@

The underlying theories that guided the selection of variables for my analysis are the two-continua model of mental health (Greenspoon \& Saklofske, 2001) and the hierarchical taxonomy of computer-mediated communication (Meier \& Reinecke, 2020).
According to the two-continua model, mental health consists of (a) psychopathology and (b) well-being.
Well-being can be differentiated into subjective and psychological well-being (Diener, Lucas, \& Oishi, 2018).
Whereas subjective well-being emphasizes hedonic aspects such as happiness and joy, psychological well-being addresses eudaimonic aspects such as fulfillment and meaning.
Subjective well-being is primarily about achieving positive affect and avoiding negative affect.
One of the most prominent indicators of well-being is life satisfaction.
In my view, because it represents a general appraisal of one's life, life satisfaction is best thought of as a meta concept that combines psychological and subjective well-being.
Notably, life satisfaction is stable and fluctuates only little, whereas it's the exact opposite for affect (Dienlin \& Johannes, 2020).
To capture well-being in this study I thus build on life satisfaction, positive affect, and negative affect.
Together, this should provide an encompassing perspective on potential media effects.

The hierarchical taxonomy of computer-mediated communication differentiates six levels of how people engage with digital technology.
First, the device (e.g., smartphone); second, the type of application (e.g., social networking site); third, the branded application (e.g., Twitter); fourth, the feature (e.g., status post); fifth, the interaction (e.g., one-to-many); and sixth, the message (e.g., content) (Meier \& Reinecke, 2020).
Whereas the first four levels focus on the communication \emph{channel}, the last two address the communication \emph{type}.
To measure social media use for the consumption of COVID-19 related news and topics, I here employ both the channel and the type of communication perspective, which together should offer a nuanced understanding of communication.

First, I investigate how well-being is affected by different types of communication affect, namely active and passive use.
Defining what constitutes active and what passive use is not always clear, and different understandings are currently discussed (Ellison, Triẹû, Schoenebeck, Brewer, \& Israni, 2020; Meier \& Reinecke, 2020).
Reading is generally considered as passive and writing as active, while there are also specific behaviors such as such liking or sharing content that fall somewhere in-between (Meier \& Krause, 2022).
In this study, I hence distinguish (a) \emph{reading} (passive), (b) \emph{posting} (active), and (c) \emph{liking and sharing} COVID-19 related posts (both active and passive).
Second, I analyze how using the most prominent branded applications affects well-being, and whether this effect changes across applications.
Branded apps are separate entities with potentially divergent effects.
Twitter might have a different effect as compared to WhatsApp because of their respective affordances.
For example, Waterloo, Baumgartner, Peter, and Valkenburg (2018) found that it's more adequate to express negative emotions on WhatsApp than on Twitter or on Instagram.
The branded applications investigated here are Facebook, Twitter, Instagram, WhatsApp, and YouTube.
Worth noting, this study is not about \emph{general} social media use during times of COVID, but on social media use \emph{focused} on COVID-19 related content.
Examples of such media use include posting thoughts about the pandemic or retweeting COVID-19 related news.

@@

Can media use be such a factor?
In advance, there doesn't seem to be a clear winner, and it's likely that both positive and negative effects cancel each other out.
Empirically, social media use on average oes not have a strong effect on well-being (Meier \& Reinecke, 2020).
According to the Differential Susceptibility to Media Effects Model (Valkenburg \& Peter, 2013), the effects of media use differ across individuals.
Whereas for some media are beneficial, for others they are harmful.
On average, however, effects are often small or negligible.
But also when focusing on individual users, social media can have both positive and negative effects on well-being (Büchi, 2021).
They can impair well-being when causing embarrassment, stress, or disinformation, and they can improve well-being when providing connectedness, information, or entertainment (Büchi, 2021).

Two of the most prominent media effect theories argue against strong negative impacts.
First, uses and gratifications theory states that people explicitly and rationally chose specific media because of their respective benefits (Katz, Blumler, \& Gurevitch, 1973).
If those benefits don't exist, they will spend their time elsewhere.
And social media offer copiuos benefits.
Most prominently, they help find relevant information, maintain and foster relationships, express one's personality, and entertain oneself (Pelletier, Krallman, Adams, \& Hancock, 2020).
The second major theory that needs to be listed here is mood management theory (Zillmann, 1988).
Users implicitly learn what media help them balance their mood and affect.
For example, when bored people use social media to entertain themselves.
Using experience sampling of well-being and logs of social media use, a study with 82 participants from Italy found that after episodes of social media use, levels of positive affect increased significantly (Marciano, Driver, Schulz, \& Camerini, 2022).

But people can also misjudge media affects and are often overly optimistic (Metzger \& Suh, 2017).
And precisely because social media have so many positive consequences, one can ask if this is not where the actual problem lies.
In other words, social media aren't problematic because they're inherently bad, but rather because they're too good.
And as with many other things, there can be too much of a good thing.
It is therefore often asked whether social media are addictive, and users sometimes express this fear themselves (Yang, Griffiths, Yan, \& Xu, 2021).
However, a recently published meta-analysis found that the two most prominent measures of addiction, the Bergen Facebook Addiction Scale and the Bergen Social Media Addiction Scale, have only small relations to well-being (Duradoni, Innocenti, \& Guazzini, 2020).
In addition, the general idea of labeling excessive social and new media use as addiction was criticized, arguing that social media represent new regular behaviors that should not be pathologized (Galer, 2018; van Rooij et al., 2018).

Because media effects can differ across situations (Valkenburg \& Peter, 2013), I now briefly focus on the effects of COVID-19 related social media use specifically.
First, one could assume a \emph{direct} negative effect on well-being, and especially on positive or negative affect, which are more volatile and fluctuating.
Dangers, inequalities, corruption---these were the headlines during the pandemic across many countries worldwide.
If one learns about such events, the initial reaction might be shock, fear, or dismay.
Consuming such news can be depressing (Dörnemann, Boenisch, Schommer, Winkelhorst, \& Wingen, 2021), perhaps even changing some general perspectives on life.
That said, because not all news was negative, and because many people showed solidarity and compassion, there was also positive and uplifting content, potentially compensating for the negative effects (Dörnemann et al., 2021).
A study with 2.057 respondents from Italy reported that during the pandemic virtual community and social connectedness even increased during the pandemic (Guazzini, Pesce, Marotta, \& Duradoni, 2022).

There could also be \emph{indirect} effects.
When browsing social media for Covid-19 related news, many users reported being captivated to such an extent that they could not stop using social media (Klein, 2021).
During the pandemic social media use was at an all-time high in the US (Statista, 2021).
Although it is most likely that moderate social media use is not detrimental (Orben, 2020), overuse, however, might be more critical, and several studies have shown more pronounced negative effects for extreme users (Przybylski \& Weinstein, 2017).
To explain, overuse could impair well-being if it replaces meaningful or functional activities such as meeting others, working, actively relaxing, or exercising.
So if a society collectively overuses social media during a pandemic, there might be potential for negative effects.
On the other hand, one can make the case that overuse might even be beneficial, especially in times of a pandemic.
Exchanging COVID-19 related messages with friends via WhatsApp might replace the in-person contact one would have otherwise, but which is literally impossible at the time.
In situations where meaningful and functional activities are prohibited, using social media to exchange about COVID-19 related topics might not be the worst idea.
Besides, given that nowadays a large number of experts, scientists, and politicians converse directly on social media, one can get first-hand high quality information on current developments.

@@

So far, there is only little empirical research on how well-being is affected by COVID-19 related social media.
In their study on the relations between media use and mental health during the pandemic, Bendau et al. (2021) found that people who used social media as a primary source of information reported on average ``significantly more unspecific anxiety and depression {[}{]} and significantly more specific COVID-19 related anxiety symptoms'' (p.~288).
Eden, Johnson, Reinecke, and Grady (2020) analyzed the media use of 425 US college students during the first wave of the pandemic, finding both positive and negative relations with well-being.
In a sample of 312 respondents collected via Amazon Mechanical Turk, Choi and Choung (2021) reported that people who used media to attain information were more lonely and less satisfied with their lives.
Stainback, Hearne, and Trieu (2020) analyzed a large-scale study with 11,537 respondents from the US and found that increased COVID-19 media consumption was related to more psychological distress.
A four-wave panel study with 384 young adults from the U.S. analyzed the effects of general digital technology use---objectively measured via screenshots of screen-time applications---on mental health, separating within- and between-person relations (Sewall, Goldstein, \& Rosen, 2021).
The results showed that digital technology did not have any significant effects on mental health (for a similar study with comparable results, see Bradley \& Howard, 2021).
Together, the literature is mixed, with a slight focus on the negative effects of social media as news use (see also Dörnemann et al., 2021; Liu \& Tong, 2020; Riehm et al., 2020).
However, note that all of these findings represent between-person relations stemming from cross-sectional data (see below).
We therefore don't know whether the differences in mental health and well-being are due to social media use or due to other third variables, such as age, health, employment, or education.

@@

Finally, also several individual studies employing advanced methods found smalls relations between social media use and well-being (Keresteš \& Štulhofer, 2020; Orben, Dienlin, \& Przybylski, 2019; Przybylski, Nguyen, Law, \& Weinstein, 2021; Schemer, Masur, Geiß, Müller, \& Schäfer, 2021).
For example, Beyens, Pouwels, van Driel, Keijsers, and Valkenburg (2021) reported that although for some users (roughly one quarter) the effects of social media use on well-being were negative, for almost the same number of users they were positive, while for the rest the effects were neutral.
This finding is aligned the Differential Susceptibility to Media Effects Model:
Although there is substantial \emph{variation} of media effects for individual users, the \emph{average} effects reported in the literature are often small (Valkenburg \& Peter, 2013).

In conclusion, in light of the theoretical considerations and empirical studies presented above, I expect that COVID-19 related communication on social media doesn't affect well-being in a meaningful or relevant way.

@@

The hypothesis explicitly states a causal effect.
In non-experimental studies, longitudinal designs can help investigate causality.
Using longitudinal designs alone, however, is not sufficient for establishing correct causal statements (Rohrer \& Murayama, 2021).
In addition, we for example also need to control for confounding third variables, and importantly also for \emph{varying} third variables.

@@

Social media use was measured for all participants on waves 1, 2, 8, 17, and 23.
Freshly recruited respondents always answered all questions on social media use.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man,mask]{apa7}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\linenumbers
\usepackage{csquotes}
\makeatletter
\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-1em}%
  {\normalfont\normalsize\bfseries\typesectitle}}

\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-\z@\relax}%
  {\normalfont\normalsize\bfseries\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
\makeatother
\usepackage{endnotes}
\setlength{\parskip}{0em}
\raggedbottom
\note{\clearpage}

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The effects of COVID-19 related social media use on well-being},
  pdflang={en-EN},
  pdfkeywords={COVID-19, well-being, social media, news use, panel study.},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{The effects of COVID-19 related social media use on well-being}
\author{Name blinded for review\textsuperscript{1}}
\date{}

@@

\abstract{%
In times of crisis such as the COVID-19 pandemic citizens need to stay informed about recent events, political decisions, or mandatory protection measures. To this end, many people use various types of media, and increasingly social media. However, because social media are particularly engaging, some find it hard to disconnect. In this preregistered study, I investigated whether using social media for COVID-19 related reasons affected psychological well-being. To answer this question I analyzed data from the Austrian Corona Panel Project, which consists of 3,485 participants. Well-being was measured at all 32 waves, and communication at six specific waves. I ran three random effects within between models, controlling for several stable and varying confounders. Results showed that the effects of COVID-19 related social media use on well-being were very small, arguably too small to matter. Fears that social media use during times of crisis critically impairs well-being are not supported.
}

@@

Some people actually couldn't stop using social media to learn about COVID-19 related news.
A new phenomenon emerged, termed ``doomscrolling'':
Users were glued to their screens and found it hard to pursue other relevant activities such as working, taking a break, or looking after their children (Klein, 2021).
It was asked whether using social media for COVID-19 related reasons is helpful, or whether it creates an additional burden on mental health (Sandstrom et al., 2021).
These concerns seem justified:
A study with 6,233 people from Germany conducted during the pandemic found that ``{[}f{]}requency, duration and diversity of media exposure were positively associated with more symptoms of depression'' (Bendau et al., 2021, p. 283).

As a result, with this study I want to build on this research and investigate whether or not COVID-19 related social media use affected well-being during the pandemic.
To this end, I analyzed a large-scale panel study from the Austrian Corona Panel Project (Kittel et al., 2020).
The panel consists of \textbf{32} waves and has an overall sample size of \textbf{3,485 participants}.
The panel study collected a large number of psychological and demographic variables.
I explicitly aimed to investigate the causal effects of COVID-19 related social media use on well-being.

@@

The underlying theories that guided the selection of variables for my analysis are the two-continua model of mental health (Greenspoon \& Saklofske, 2001) and the hierarchical taxonomy of computer-mediated communication (Meier \& Reinecke, 2020).
According to the two-continua model, mental health consists of (a) psychopathology and (b) well-being.
Well-being can be differentiated into subjective and psychological well-being (Diener et al., 2018).
Whereas subjective well-being emphasizes hedonic aspects such as happiness and joy, psychological well-being addresses eudaimonic aspects such as fulfillment and meaning.
Subjective well-being is primarily about achieving positive affect and avoiding negative affect.
One of the most prominent indicators of well-being is life satisfaction.
In my view, because it represents a general appraisal of one's life, life satisfaction is best thought of as a meta concept that combines psychological and subjective well-being.
Notably, life satisfaction is stable and fluctuates only little, whereas it's the exact opposite for affect (Dienlin \& Johannes, 2020).
To capture well-being in this study I thus build on life satisfaction, positive affect, and negative affect.
Together, this should provide an encompassing perspective on potential media effects.

The hierarchical taxonomy of computer-mediated communication differentiates six levels of how people engage with digital technology.
First, the device (e.g., smartphone); second, the type of application (e.g., social networking site); third, the branded application (e.g., Twitter); fourth, the feature (e.g., status post); fifth, the interaction (e.g., one-to-many); and sixth, the message (e.g., content) (Meier \& Reinecke, 2020).
Whereas the first four levels focus on the communication \emph{channel}, the last two address the communication \emph{type}.
\textbf{Differentiating different communication channels and communication types is important, because the effects on well-being likely differ across communication channels and communication types.}
\textbf{Whereas active social media use such as chatting is routinely linked to improved well-being, passive use such as reading is considered more negative (Dienlin \& Johannes, 2020).}
Similarly, branded apps are separate entities with potentially divergent effects.
For example, Waterloo et al. (2018) found that it's more adequate to express negative emotions on WhatsApp than on Twitter or on Instagram.
\textbf{Especially during a pandemic, it makes sense to analyze if users engage with COVID-19 related content on Instagram, where communication is more positive, or on Facebook, where communication is more critical.}

\textbf{In this study, to measure the effects of social media use focused on COVID-19 related news and topics, I adopt both the channel and the type of communication perspective.}
\textbf{Together, this should offer a nuanced and comprehensive understanding of communication.}
First, I investigate how well-being is affected by different types of communication affect, namely active and passive use.
Defining what constitutes active and what passive use is not always clear, and different understandings are currently discussed (Ellison et al., 2020; Meier \& Reinecke, 2020).
Reading is generally considered as passive and writing as active, while there are also specific behaviors such as such liking or sharing content that fall somewhere in-between (Meier \& Krause, 2022).
In this study, I hence distinguish (a) \emph{reading} (passive), (b) \emph{posting} (active), and (c) \emph{liking and sharing} COVID-19 related posts (both active and passive).
Second, I analyze how using the most prominent branded applications affects well-being, and whether this effect changes across applications.
The branded applications investigated here are Facebook, Twitter, Instagram, WhatsApp, and YouTube---which were, at the time of writing, the most relevant social media apps.

\textbf{Worth noting, this study is not about \emph{general} social media use during times of COVID, but on social media use \emph{focused} on COVID-19 related content.}
\textbf{For example, posting thoughts about the pandemic, reading posts and comments, or retweeting COVID-19 related news.}

@@

Can media use be such a factor?
In advance, there doesn't seem to be a clear winner, and it's likely that both positive and negative effects cancel each other out.
Empirically, social media use on average does not have a strong effect on well-being (Meier \& Reinecke, 2020).
According to the Differential Susceptibility to Media Effects Model (Valkenburg \& Peter, 2013), the effects of media use differ across individuals \textbf{and content}.
Whereas for some media are beneficial, for others they are harmful.
\textbf{Whereas some content provides opportunities (education, advice), other content creates risks (misinformation, hate). (Livingstone et al., 2018).}
Social media can impair well-being when causing embarrassment, stress, or disinformation, and they can improve well-being when providing connectedness, information, or entertainment (Büchi, 2021).
On average, however, effects are often small or negligible.

\textbf{Two of the most prominent media effect theories argue (indirectly) against strong average negative impacts.}
\textbf{According to mood management theory (Zillmann, 1988), using media substantially affects people's moods.}
\textbf{Effects can be stimulating or overwhelming, relaxing or boring.}
\textbf{After some time, users implicitly learn what media help them balance their mood and affect according to their own situational needs (Zillmann, 1988).}
\textbf{Those media that eventually become part of one's media repertoire are hence, on average, beneficial for users and their moods.}
\textbf{Using experience sampling of well-being and logs of social media use, a study with 82 participants from Italy found that after episodes of social media use, levels of positive affect increased significantly (Marciano et al., 2022).}

\textbf{While mood management theory considers media use mainly driven by implicit learning experiences, uses and gratifications theory upholds that the process is more explicit and rational (Katz et al., 1973).}
\textbf{Users select those media that they expect to have a desired effect, for example on mood, knowledge, or entertainment.}
\textbf{If those beneficial media effects do not exist or if they are not expected, people will spend their time elsewhere.}
\textbf{And social media offer several beneficial effects, explaining why they are used that much.}
\textbf{They help find relevant information, maintain and foster relationships, express one's personality, and entertain oneself (Pelletier et al., 2020).}
\textbf{In conclusion, because people spend so much time on social media consuming COVID-19 related content, according to both mood management theory and uses and gratifications theory we wouldn't expect to find strong average negative effects.}

But people can also misjudge media affects and are often overly optimistic (Metzger \& Suh, 2017).
And precisely because social media have so many positive consequences, one can ask if this is not where the actual problem lies.
In other words, social media aren't problematic because they're inherently bad, but rather because they're too good.
And as with many other things, there can be too much of a good thing.
It is therefore often asked whether social media are addictive, and users sometimes express this fear themselves (Yang et al., 2021).
However, a recently published meta-analysis found that the two most prominent measures of addiction, the Bergen Facebook Addiction Scale and the Bergen Social Media Addiction Scale, have only small relations to well-being (Duradoni et al., 2020).
In addition, the general idea of labeling excessive social and new media use as addiction was criticized, arguing that social media represent new regular behaviors that should not be pathologized (Galer, 2018; van Rooij et al., 2018).

\textbf{Because media effects can differ across users, situations, and content (Livingstone et al., 2018; Valkenburg \& Peter, 2013), I now briefly focus on the effects of COVID-19 related social media use specifically.}
First, one could assume a \emph{direct} negative effect on well-being, and especially on positive or negative affect, which are more volatile and fluctuating.
Dangers, inequalities, corruption---these were the headlines during the pandemic across many countries worldwide.
If one learns about such events, the initial reaction might be shock, fear, or dismay.
Consuming such news can be depressing (Dörnemann et al., 2021), perhaps even changing some general perspectives on life.
That said, because not all news was negative, and because many people showed solidarity and compassion, there was also positive and uplifting content, potentially compensating for the negative effects (Dörnemann et al., 2021).
A study with 2.057 respondents from Italy reported that during the pandemic virtual community and social connectedness even increased during the pandemic (Guazzini et al., 2022).
\textbf{In Finland, in a sample of 735 participants, levels of loneliness did not decrease during the pandemic, and people who engaged more on social media showed less loneliness (Latikka et al., 2022).}

There could also be \emph{indirect} effects.
When browsing social media for Covid-19 related news, many users reported being captivated to such an extent that they could not stop using social media (Klein, 2021).
During the pandemic social media use was at an all-time high in the US (Statista, 2021).
Although it is most likely that moderate social media use is not detrimental (Orben, 2020), overuse, however, might be more critical, and several studies have shown more pronounced negative effects for extreme users (Przybylski \& Weinstein, 2017).
To explain, overuse could impair well-being if it replaces meaningful or functional activities such as meeting others, working, actively relaxing, or exercising.
\textbf{Another potentially negative mechanism at play are problematic social comparison processes. During the pandemic, several users shared how they successfully dealt with challenges such as physical distancing. In a study with 1131 residents from Wuhan in China (Yue et al., 2022), people who spent more time in quarantine also spent more time on social media. Those, who spent more time on social media also engaged in more upward social comparison, which was related to increased levels of stress.}

On the other hand, one can make the case that using social media for COVID-19 related reasons might even be beneficial, especially in times of a pandemic.
Exchanging COVID-19 related messages with friends via WhatsApp might replace the in-person contact one would have otherwise, but which is literally impossible at the time.
In situations where meaningful and functional activities are prohibited, using social media to exchange about COVID-19 related topics might not be the worst idea.
Besides, given that nowadays a large number of experts, scientists, and politicians converse directly on social media, one can get first-hand high quality information on current developments.

@@

So far, there is only little empirical research on how well-being is affected by COVID-19 related social media.
In their study on the relations between media use and mental health during the pandemic, Bendau et al. (2021) found that people who used social media as a primary source of information reported on average ``significantly more unspecific anxiety and depression {[}{]} and significantly more specific COVID-19 related anxiety symptoms'' (p.~288).
Eden et al. (2020) analyzed the media use of 425 US college students during the first wave of the pandemic, finding both positive and negative relations with well-being.
In a sample of 312 respondents collected via Amazon Mechanical Turk, Choi and Choung (2021) reported that people who used media to attain information were more lonely and less satisfied with their lives.
Stainback et al. (2020) analyzed a large-scale study with 11,537 respondents from the US and found that increased COVID-19 media consumption was related to more psychological distress.
A four-wave panel study with 384 young adults from the U.S. analyzed the effects of general digital technology use---objectively measured via screenshots of screen-time applications---on mental health, separating within- and between-person relations (Sewall et al., 2021).
The results showed that digital technology did not have any significant effects on mental health (for a similar study with comparable results, see Bradley \& Howard, 2021).
Together, the literature is mixed, with a slight focus on the negative effects of social media as news use (see also Dörnemann et al., 2021; Liu \& Tong, 2020; Riehm et al., 2020).
However, note that all of these findings represent between-person relations stemming from cross-sectional data (see below).
We therefore don't know whether the differences in mental health and well-being are due to social media use or due to other third variables, such as age, health, employment, or education.

@@

Finally, also several individual studies employing advanced methods found smalls relations between social media use and well-being (Keresteš \& Štulhofer, 2020; Orben et al., 2019; Przybylski et al., 2021; Schemer et al., 2021).
For example, Beyens et al. (2021) reported that although for some users (roughly one quarter) the effects of social media use on well-being were negative, for almost the same number of users they were positive, while for the rest the effects were neutral.
This finding is aligned the Differential Susceptibility to Media Effects Model:
Although there is substantial \emph{variation} of media effects for individual users, the \emph{average} effects reported in the literature are often small (Valkenburg \& Peter, 2013).

In conclusion, in light of the theoretical considerations and empirical studies presented above, I expect that COVID-19 related communication on social media doesn't affect well-being in a meaningful or relevant way.
\textbf{This general hypothesis will be analyzed specifically for the communication types of (a) time spent reading, (b) liking and sharing, and (c) actively posting COVID-19 related content.}
\textbf{In addition, I will analyze how well-being is influenced by spending time on five prominent social media apps, including (a) Facebook, (b) Instagram, (c) Twitter, (d) WhatsApp, and (e) YouTube.}
\textbf{Three different well-being indicators will be differentiated: life satisfaction, positive affect, and negative affect.}

@@

The hypothesis explicitly states a causal effect.
In non-experimental studies, longitudinal designs can help investigate causality.
Using longitudinal designs alone, however, is not sufficient for establishing correct causal statements (Rohrer \& Murayama, 2021).
In addition, we for example also need to control for confounding third variables.
\textbf{Importantly, when analyzing longitudinal (within-person) relationships and effects, it is important to control for \emph{varying} third variables.}
\textbf{Non-varying third variables only help control non-varying (between-person) relations.}

@@

Social media use was measured for all participants on waves 1, 2, 8, 17, 23, \textbf{and 28}.
Freshly recruited respondents always answered all questions on social media use.
